%
%   Timed Mobile Systems
%   Andrew Hughes and Mike Stannett
%   Basic Equivalence Theory
%
\documentclass[orivec,envcountsame]{llncs}
\usepackage{verbatim}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{amsfonts, amssymb, amsmath}

\usepackage{bussproofs}

\newcommand{\Exhibits}[1]{\mathrel{\downarrow_{#1}}}
\newcommand{\ExhibitsA}{\Exhibits{\mathscr{A}}}
\newcommand{\ExhibitsC}{\Exhibits{\mathscr{C}}}
\newcommand{\ExhibitsE}{\Exhibits{\mathscr{E}}}

\newcommand{\Reveals}[1]{\mathrel{\Downarrow_{#1}}}
\newcommand{\RevealsA}{\Reveals{\mathscr{A}}}
\newcommand{\RevealsC}{\Reveals{\mathscr{C}}}
\newcommand{\RevealsE}{\Reveals{\mathscr{E}}}

%\newcommand{\Invokes}[1]{\mathrel{\Downarrow_{#1}}}
%\newcommand{\InvokesA}{\Invokes{\mathscr{A}}}
%\newcommand{\InvokesC}{\Invokes{\mathscr{C}}}
%\newcommand{\InvokesE}{\Invokes{\mathscr{E}}}

\newcommand{\Eq}{\approx}
\newcommand{\EqA}{\mathrel{\Eq_{\mathscr{A}}}}
\newcommand{\EqC}{\mathrel{\Eq_{\mathscr{C}}}}
\newcommand{\EqE}{\mathrel{\Eq_{\mathscr{E}}}}

\newcommand{\Means}{\mathrel{~=_{\mathrm{def}}~}}
\newcommand{\Does}[1]{\xderives{#1}}
\newcommand{\DoesTaus}{\mathrel{\Rightarrow}}
\newcommand{\Theorem}{\vdash}
\newcommand{\OfType}{\mathrel{:}}

\input{macros}
\input{mpsmacros}

\begin{document}


\section{Equivalence Theory}
\label{sec:equivalence-theory}

What does it mean for two environs $\locv{m}{P}{B}{\vec\rho}$ and
$\locv{n}{Q}{C}{\vec\sigma}$ to be equivalent? We will adopt the standard Morris-style  \cite{Mor68} (or \emph{may-testing} \cite{DNH84}) approach used to characterise
both equivalence of Mobile Ambients \cite{GC99} and barbed congruence in the
$\pi$-calculus \cite{SW01}. Having defined equivalence, we can prove various
safety properties. In particular, we show that $P \Eq Q$ implies that $P$ and
$Q$ have the same type relative to our type system.

\subsection{Immediate Observability}
\label{sec:immediate-observability}

We first define three predicates $\expr \ExhibitsA \alpha$, $\expr \ExhibitsC
\sigma$ and $\expr \ExhibitsE n$, describing the immediate observability of
actions and clocks, and the immediate top-level placement of environs,
respectively. For each predicate $\Exhibits{\bullet}$ we then define a
corresponding contextual equivalence $\Eq_{\bullet}$. Finally, we define
equivalence $\Eq$ for TNT processes to be the conjunction of these three
equivalences. We use the word \emph{symbol} to mean any action, clock or environ
name.

\subsubsection{Actions.}
The notion of immediate observability of an action needs to take account of possible timeouts. For example, the action $a$ can occur immediately in both $a\pref\nil$ and $\timeout{a\pref\nil}{\sigma}{b\pref\nil}$, while the action $b$ can occur immediately in neither. We can capture the required behaviour quite simply, however, by appealing to our semantics.

\begin{definition}
We say that $\expr$ \emph{exhibits} the action $\alpha$, and write $\expr \ExhibitsA \alpha$, provided $\expr \Does{\alpha}$.
\end{definition}


\subsubsection{Clocks.} 

We consider a clock to be immediately observable if it can tick immediately.

\begin{definition}
We say that $\expr$ \emph{exhibits} the clock $\sigma$, and write $\expr \ExhibitsC \sigma$, provided $\expr \Does{\sigma}$.
\end{definition}

\subsubsection{Environs} 

Following Gordon and Cardelli \cite{GC99} we say that $\expr$ \emph{exhibits}
the environ $n$ whenever $n$ is the name of a `top-level' environ. However, our
definition again needs to take account of potential time-outs. For example, $n$
is exhibited in both $\loc{n}{P}{B}{\sigma}$ and
$\timeout{\loc{n}{P}{B}{\sigma}}{\sigma}{\nil}$. In addition, the presence of time-out operators means that restrictions can occur at several locations in the expression, without affecting the environ's visibility.

\begin{definition}
We say that $\expr$ \emph{exhibits} the environ $n$, and write $\expr \ExhibitsE
n$, provided $\expr$ is of one of the forms 
\begin{enumerate}
\item 
    $\locv{n}{\exprb}{B}{\vec{\sigma}}$;
\item
    $\timeout{ \expr' }{\rho}{\exprb}$ where $\expr' \ExhibitsE n$;
\item
    $\stimeout{ \expr' }{\rho}{\exprb}$ where $\expr' \ExhibitsE n$; or
\item
    $\expr' \hide{\vec{\sigma}}$ where $\expr' \ExhibitsE n$.
\item
    $\expr' \res{ \vec{n} }$ where $n \not\in \vec{n}$ and $\expr' \ExhibitsE n$.
\end{enumerate}
\end{definition}


\subsection{Eventual Observability}
\label{sec:eventual-observability}

A symbol $x$ is \emph{eventually observable} (we shall sometimes say it is \emph{revealed}) in an expression $\expr$ precisely when $x$ is the first \emph{relevant} action of the appropriate class. As usual, we write $\DoesTaus$ for the reflexive transitive closure of $\Does{\tau}$.


\subsubsection{Actions.}

There is only one unobservable action, $\tau$. This does not mean, however, that $\tau$ cannot be revealed. Rather, we don't care about \emph{additional} occurrences of $\tau$.

\begin{definition}
We say that $\expr$ \emph{reveals} the action $\alpha$, and write $\expr \RevealsA \alpha$, provided there is some $\expr'$ such that $\expr \DoesTaus \expr' \ExhibitsA \alpha$.
\end{definition}

\subsubsection{Clocks.}

Clocks are used to delimit activities, so we should allow normal actions to proceed between clock ticks without prejudicing the subsequent tick's right to be considered eventually observed.

\begin{definition}
We say that $\expr$ \emph{reveals} the clock $\sigma$, and write $\expr \RevealsC \sigma$, provided there is some $\expr'$ and some $\vec{s} \in \left( \act \cup \{ \tin, \tout, \topen \}\right)^*$ such that $\expr \Does{\vec{s}} \expr' \ExhibitsC \sigma$.
\end{definition}

Since clocks are automatically hidden by enclosure in an environ, and cannot be introduced syntactically except within time-out, the following characterisation is straightforward.

\begin{proposition}
If $\expr \RevealsC \sigma$, then $\expr$ is of one of the forms
\begin{enumerate}
\item
    $\timeout{\exprb}{\sigma}{\exprb'}$;
\item
    $\stimeout{\exprb}{\sigma}{\exprb'}$;
\item
    $\timeout{\expr'}{\sigma}{\exprb'}$ where $\expr' \ExhibitsC \sigma$;
\item
    $\stimeout{\expr'}{\sigma}{\exprb'}$ where $\expr' \ExhibitsC \sigma$;
\item
    $\expr' \hide{\vec{\sigma}}$ where $\sigma \not\in \vec{\sigma}$ and $\expr' \ExhibitsC n$.
\item
    $\expr' \res{ \vec{n} }$ where $\expr' \ExhibitsC \sigma$.
\end{enumerate}
\end{proposition}
\begin{proof}
Straightforward. \textbf{Am checking.}
\qed \end{proof}



\subsubsection{Environs.}

An environ is eventually observed provided it can be brought to top level by some sequence of mobility-based reductions.

\begin{definition}
We say that $\expr$ \emph{reveals} the environ $n$, and write $\expr \RevealsE n$, provided there is some $\expr'$ and some $\vec{\ambop} \in \{\tin, \tout, \topen\}^*$ such that $\expr \Does{\vec{\ambop}} \expr' \ExhibitsE n$.
\end{definition}



\subsection{Contextual Congruence}
\label{sec:contextual-congruence}

As usual, a \emph{context} $C$ is an expression with one or more `holes'. We write $[\,]$ for a hole, and $C[\expr]$ for the expression obtained by filling $C$'s holes with copies of $\expr$ (this may result in free names and variables in $\expr$ becoming bound).

\begin{definition}
Let $C$ range over arbitrary contexts. Two expressions are
\begin{enumerate}
\item
    \emph{action equivalent}, written $\expr \EqA \exprb$, provided they reveal the same actions in every context: $C[\expr] \RevealsA \alpha \Leftrightarrow C[\exprb] \RevealsA \alpha$;
\item
    \emph{clock equivalent}, written $\expr \EqC \exprb$, provided they reveal the same clocks in every context: $C[\expr] \RevealsC \sigma \Leftrightarrow C[\exprb] \RevealsC \sigma$;

\item
    \emph{environ equivalent}, written $\expr \EqE \exprb$, provided they reveal the same environs in every context: $C[\expr] \RevealsE n \Leftrightarrow C[\exprb] \RevealsE n$.
\end{enumerate}
\end{definition}

Finally, full equivalence of TNT expressions is defined to be the conjunction of action, clock and mobility equivalence.
\begin{definition}[Expression Equivalence]
Two expressions $\expr$ and $\exprb$ are (contextually) \emph{equivalent}, written $\expr \Eq \exprb$, provided they are action-, clock- and environ-equivalent. That is,
\[
    \expr \Eq \exprb
    \Means
    \left( \expr \EqA \exprb \right)
    \land
    \left( \expr \EqC \exprb \right)
    \land 
    \left( \expr \EqE \exprb \right)
\]
\end{definition}

\begin{lemma}
Contextual equivalence is a congruence.
\end{lemma}
\begin{proof}
Straightforward \textbf{Am checking.}
\qed \end{proof}

Although we have defined contextual congruence without reference to the type system, it is important to show that they respect one another. The following theorem illustrates the coherence of our definitions.

\begin{theorem}
If $\expr \Eq \exprb$ and $\Theorem \expr \OfType g$, then $\Theorem \exprb \OfType g$.
\end{theorem}
\begin{proof}
\textbf{Hmmmm.} 
\qed \end{proof}

\bibliographystyle{splncs}
\bibliography{literature}


\end{document}
